---
title: "Konza Vegetation Analysis"
author: "Naupaka Zimmerman"
format: html
execute: 
  echo: false
  warning: false
  message: false
---

## Introduction

This analysis will look at vegetation data from the Konza National Ecological Observatory Network (NEON) site located near Manhattan, KS. It will use tidyverse packages to wrangle and visualize some of the data.

## Loading the data

The data are provided in csv files. We'll use the tidyverse package `readr` to load it into R.

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")
library("ggthemes")
library("knitr")

# read in the data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z.csv")
```

## Data summarizing with `dplyr`

We'll next use the dplyr package to summarize some different aspects of the data.

```{r}
#| label: summarize-with-dplyr

# first summary to make will be a count of the number of observations in each family
konza_data %>%
  group_by(family) %>%
  tally() %>%
  arrange(desc(n)) %>%
  kable()
```

This analysis showed that Poaceae and Asteraceae are among the most common family found in these plots.

```{r}
#| label: mean-pcnt-cover-by-family

# the goal is to calculate the mean percent cover of each family
konza_data %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover)) %>%
  kable()
```

This analysis showed that the largest percent cover is the `NA` taxa, which means it likely wasn't able to be identified, but the highest mean percent cover at 16% across the plots was Fagaceae.

```{r}
#| label: filter-by-elevation

# the goal is to calculate the mean percent cover of each family
konza_data %>%
  filter(elevation > 390) %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover)) %>%
  kable()
```

Next we'll learn to use `mutate()`, which is a function to create a new column based on contents from an existing column.

```{r}
#| label: use-mutate

# select pulls out the columsn you want
# mutate creates a new column with whatever content you want
# which can include manipulation of an existing column
konza_data %>%
  select(scientificName, family, percentCover) %>%
  mutate(propCover = percentCover / 100) %>%
  top_n(20) %>%
  kable()
```

Next up, we'll use `ggplot2` to make some figures!

```{r}
#| label: histogram-percent-cover

konza_data %>%
  filter(!is.na(family)) %>%
  ggplot(aes(x = percentCover)) +
  geom_histogram() +
  labs(title = "Histogram of percent cover, all species",
       x = "Percent Cover") +
  theme_bw()
```

## Next steps

1. Follow up on the really high percent cover values. Maybe say > 25%?

Looks like it's mostly soil and dead plant matter.


